---
title: "Introduction to R - R Ladies Vienna"
author: "Pia Pfeiffer and Patricia Puchhammer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_chunk$set(echo = TRUE)
```

## What is R? <https://www.r-project.org/>

> R is a programming language and software environment specifically designed for statistical computing and graphics. It provides a wide variety of statistical and graphical techniques, making it a preferred tool for data analysis and visualization among statisticians, data scientists, researchers, and analysts [...] across various industries and domains.
>
> `r tufte::quote_footer('--- OpenAI. (2024). ChatGPT (3.5) [Large language model] https://chat.openai.com')`

-   Firms like Google, Pfizer, Bank of America, Amazon and Microsoft use R for their statistical analyses.

-   R is open source! freely available for anyone to use, modify, and distribute, active community

### Statistics

-   Most classical statistics and much of the latest methodology is available for use with R

-   A few of these are built into the base R environment, but many are supplied as packages available through the *The Comprehensive R Archive Network* (CRAN, <https://CRAN.R-project.org>)

-   Visualization: publication-quality plots and charts

-   Reproducibility: R promotes reproducible research by enabling users to document their analyses in a script format.

### Workshop Content

-   Intro to RStudio
-   Basic data types and structures
-   Import and export of data sets
-   Basic data processing
-   Basic data visualization and `ggplot2`
-   Loops, functions, and conditional statements

## Getting Started

### Getting help

R is a collaborative project with many contributors. Type 'contributors()' for more information and 'citation()' on how to cite R or R packages in publications. Type 'demo()' for some demos, 'help()' or **'?'** for on-line help, or 'help.start()' for an HTML browser interface to help. Type 'q()' to quit R.

-   Beginner's guide: <https://support.posit.co/hc/en-us/articles/201141096-Getting-Started-with-R>
-   Manual: <https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf>
-   R Cookbook: <https://rc2e.com/>
-   R for Data Science: <https://r4ds.hadley.nz/>

### Installation

-   Downloading and installing R. <https://cran.r-project.org/>
-   Downloading and installing RStudio. <https://posit.co/downloads/>
-   Download code for R Introduction from <https://github.com/patriciapuch>/

### RStudio

-   RStudio is an integrated development environment (IDE) for R.

-   It includes a console, syntax-highlighting editor that supports direct code execution, as well as tools for plotting, history, debugging and workspace management.

-   RStudio is available in open source and commercial editions.

-   RStudio runs on the desktop (Windows, Mac, and Linux) or in a browser connected to RStudio Server or RStudio Server Pro (Debian/Ubuntu, RedHat/CentOS, and SUSE Linux).

-   Introduction to RStudio interface.

### Basic commands and arithmetic operations

-   Using R as a calculator

```{r basic_commands}
3 + 4
3 - 7
2 * 20
4/2
10^3
sqrt(10)
```

-   Comparisons

```{r comparisons}
3 < 5
3 > 5
3 <= 5 
3 >= 5
3 == 5     # Attention: double equal sign
3 != 5
```

### Variables

-   Assignment

```{r assignment}
x <- 3
y = 4
5 -> z # works, but not recommended
```

## Data Structures

Creating and manipulating data structures.

-   Atomic vectors (elements are same type)

```{r atomic_vectors}
pyth_triple <- c(x, y, z)
?c
# indexing with boolean (condition)
pyth_triple[pyth_triple >= 4]
```

-   Sequences (generate regular sequences)

```{r sequences}
seq_incr <- seq(1, 5, by = 1)
seq_decr <- 10^seq(4, 0, by = -1)
```

-   Factors (for categorical data, can be ordered, levels are unique values)

```{r factors}
x1 <- c("Dec", "Apr", "Jan", "Mar")
x2 <- c("Dec", "Apr", "Jam", "Mar")
sort(x1)
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
y1 <- factor(x1, levels = month_levels)
y2 <- factor(x2, levels = month_levels)
sort(y1)
```

-   Matrices (2-dimensional generalization of vectors)

```{r matrices}
mat <- matrix(0, nrow = 5, ncol = 5)
diag(mat) <- c(seq(1, 2, by = 1), seq(4, 2, by = -1))
mat
```

-   Data Frames (matrix-like structures, can handle different data types)

```{r data_frames}
v_char <- paste0("char", seq(1, 10, 1))
v_num <- seq(0.01, 0.1, length.out = 10)
v_bool <- c(rep(TRUE, 4), rep(FALSE, 6))
df <- data.frame(col1 = v_char,
                 col2 = v_num,
                 col3 = v_bool)
df$col1[df$col3] #indexing
df$col2[df$col2 > 0.05]
```

-   Lists (general form of vector, elements can be of different type)

```{r lists}
my_list <- list(matrix = mat, 
                vector = pyth_triple, 
                fact = y2, 
                bool = TRUE)
```

## Packages

-   CRAN: <https://cran.r-project.org/>

-   Bioconductor: <https://www.bioconductor.org/>

-   find packages that might be useful for your field: <https://cran.r-project.org/web/views/>

-   popular packages for Data Science are combined in 'tidyverse' <https://www.tidyverse.org/> (**ggplot2**, dplyr, tidyr, readr, stringr, ...)

-   Installation (CRAN)

```{r installation, eval = FALSE}
install.packages("tidyverse")
```

-   get help with packages and functions

```{r load_packages}
library(tidyverse)
citation("tidyverse")
```

```{r help_packages}
help(package = "tidyverse")
```

## Data Import and Export

-   Importing data from CSV, Excel, text files.

```{r data_import}
?read.table

# getwd()
# setwd("C:/Users/puchhammer/Documents/Konferenzen Workshops/R_Ladies2025_IntroToR/intro-to-R-main/")

path1 <- file.path("iris", "iris_1.csv")
path2 <- file.path("iris", "iris_2.csv")
path3 <- file.path("iris", "iris_3.txt")
path4 <- file.path("iris", "iris_4.txt")

iris1 <- read.csv(path1, header = T)
iris2 <- read.csv2(path2, dec = ".")
iris3 <- read.table(path3, header = T)
iris4 <- read.table(path4)

head(iris1) # look at data


```

-   Data cleaning and preprocessing.
-   Useful package `janitor`: <https://cran.r-project.org/web/packages/janitor/index.html>

```{r data_proc}
?iris

# selecting and naming columns
iris_new <- iris1[,2:6]
colnames(iris_new) <- c("sepal_l", "sepal_w", "petal_l", "petal_w", "species")

# missing values
iris_new[5,2] <- NA
mean(iris_new[,2], na.rm = TRUE)
iris_new <- na.omit(iris_new)
mean(iris_new[,2])

# modifying columns
iris_new$total_l <- iris_new$sepal_l + iris_new$petal_l
iris_new$total_w <- iris_new$sepal_w + iris_new$petal_w
iris_new$species <- factor(iris_new$species, levels = c( "setosa", "versicolor", "virginica"))
head(iris_new)
```

-   Exporting data.

```{r data_exp}
?write.table

write.csv(iris_new, file = file.path("iris", "iris_mod.csv"))

save(iris_new, file = file.path("iris", "iris_mod.Rdata"))
load(file = file.path("iris", "iris_mod.Rdata"))
```

-   package for data wrangling: 'dplyr' <https://dplyr.tidyverse.org/>

## Data Visualization

-   Creating and modifying plots with base R bar plots, histograms.

```{r basic_plots}
plot(iris_new$sepal_l, iris_new$petal_l, 
     type = "p", 
     lty = 1, 
     col = iris_new$species,
     xlab = "Sepal Length", ylab = "Petal Length")
points(iris_new$sepal_l[100], iris_new$petal_l[100], 
       pch = 19,
       col = iris_new$species[100])
legend("bottomright", fill = c("black", "red", "green"), legend = levels(iris_new$species))
abline(v = iris_new$sepal_l[100], lty = 2)
abline(h = iris_new$petal_l[100], lty = 3)
abline(coef = c(0.6, 0.75), col = "green")
```

```{r histogram}
hist(iris_new$petal_w, 
     breaks = 10,
     main = "Example for Histogram",
     xlab = "Petal Width",
     col = "lightblue", lty = 2)
abline(v = 0.7, lty = 1, col = "red")
```

-   multivariate: pairs

```{r}
pairs(iris_new[,1:2], 
      col = iris_new$species, 
      pch = as.numeric(iris_new$species),
      labels = c("Sepal Length", "Sepal Width"))
```

-   Export of plots

```{r plots_export}

```

-   ggplot2 <https://ggplot2.tidyverse.org/>

```{r ggplots}
library(ggplot2)
ggplot(data = iris_new,
       aes(x = total_w, 
           y = total_l, 
           groups = species,
           col = species)) + 
  geom_point() +
  geom_line(aes(linetype = species)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Total Width") + ylab("Total Length") +
  labs(col = "Species", linetype = "Species") +
  scale_color_brewer(palette = "Set1") + 
  theme_light()

? scale_color_brewer 
? theme_light
```

## Programming

-   if else

```{r if_else}
if (3 > 5) {
  print("That's true!") 
  } else {
  print("That's false")
  }
```

-   for, while

```{r loops}
for (i in 1:3){
  print(paste("count:", i))
}

j <- 0
while (j < 4){
  print(j)
  j <- j + 1
}
```

-   functions

```{r functions}
my_plot <- function(x, y, color, linetype) {
  plot(x, y, type = "l", col = color, lty = linetype)
}

my_plot(seq(1,10,1), seq(1,10,1)^2, 5, 2)
```

## R Markdown

-   Reporting and reproducible research
-   Introduction: <https://rmarkdown.rstudio.com/>
